### 2011-06-05 docs/pubs/EIM-2011/abstract.tex
Trivial spelling and grammar fixes.

\begin{abstract}
  Effective discovery and integration of ecological data within data
  management systems requires rich semantic information that can
  describe and relate the types of information contained within
  disparate data sets. Within the Semtools project, we have developed
  approaches for expressing and representing semantic annotations of
  data sets for supplementing attribute and data-level metadata with
  terms drawn from domain-specific ontologies. Annotations provide a
  formal mechansism that can be used together with reasoning systems
  to enhance existing data discovery and integration approaches.  In
  this paper we describe extensions to the Ecological Metadata
  Language (EML) and associated tools for storing and using semantic
  annotations.  Specifically, we describe new user interface
  components implemented within the Morpho EML metadata editor for
  capturing user-supplied semantic annotations, extensions to the
  Metacat system for storing and accessing annotations and
  corresponding OWL-DL ontologies, and a new API within Metacat that
  uses annotation metadata to provide concept-based search and
  integration of data sets.
%   The Data Manager paper from the last EIM included semantic
%   foreshadowing in the 'future directions' section and this is a
%   logical continuation of that work. Previously, we saw that the DM
%   library enabled data integration tasks provided we knew exactly
%   which data attributes from exactly which datasets were "compatible"
%   with each other. Here "compatible" means not simply structurally
%   similar (e.g. decimal numbers) but also conceptually similar
%   (e.g. ocean water temperature in C). By including rich semantic
%   annotations we initially show simple integration capabilities where
%   the subject domain of our data corpus is relatively constrained
%   (i.e. SBC-LTER) and the observational models represented are
%   minimally variable (i.e. context excluded). Semantically-enahnced
%   query capabilities facilitate this 'smart integration' in multiple
%   ways.
%   First, the corpus is refined through compound concept-based
%   queries that exercise OBOE-compatible ontological subsumption
%   hierarchies to locate roughly equivalent attributes across
%   datasets. Additional query constraints are defined in relation to
%   the precise observational model formally expressed in OBOE. Finally,
%   the actual data values are interrogated so that only those that fall
%   in the desired range are returned. From this resultant subset of
%   disparate data objects we create a single synthetic data product
%   that represents a 'materialization' of the observations. Technical
%   discussion will include high-level descriptions of the
%   infrastructure and tools employed with focus on our development in
%   the rapidly evolving semantic frontier where scalability and best
%   practices are particularly germane and equally elusive.
\end{abstract}

>>>>

\begin{abstract}
  Effective discovery and integration of ecological data within data
  management systems requires rich semantic information that can
  describe and relate the types of information contained within
  disparate data sets. Within the Semtools project, we have developed
  approaches for expressing and representing semantic annotations of
  data sets for supplementing attribute and data-level metadata with
  terms drawn from domain-specific ontologies. Annotations provide a
  formal mechansism that can be used together with reasoning systems
  to enhance existing data discovery and integration approaches.  In
  this paper we describe extensions to the Ecological Metadata
  Language (EML) and associated tools for storing and using semantic
  annotations.  Specifically, we describe new user interface
  components implemented within the Morpho metadata editor for
  capturing user-supplied semantic annotations, extensions to the
  Metacat system for storing and accessing annotations and
  corresponding OWL-DL ontologies, and a new API within Metacat that
  uses annotation metadata to provide concept-based search and
  integration of data sets.
%   The Data Manager paper from the last EIM included semantic
%   foreshadowing in the 'future directions' section and this is a
%   logical continuation of that work. Previously, we saw that the DM
%   library enabled data integration tasks provided we knew exactly
%   which data attributes from exactly which datasets were "compatible"
%   with each other. Here "compatible" means not simply structurally
%   similar (e.g. decimal numbers) but also conceptually similar
%   (e.g. ocean water temperature in C). By including rich semantic
%   annotations we initially show simple integration capabilities where
%   the subject domain of our data corpus is relatively constrained
%   (i.e. SBC-LTER) and the observational models represented are
%   minimally variable (i.e. context excluded). Semantically-enahnced
%   query capabilities facilitate this 'smart integration' in multiple
%   ways.
%   First, the corpus is refined through compound concept-based
%   queries that exercise OBOE-compatible ontological subsumption
%   hierarchies to locate roughly equivalent attributes across
%   datasets. Additional query constraints are defined in relation to
%   the precise observational model formally expressed in OBOE. Finally,
%   the actual data values are interrogated so that only those that fall
%   in the desired range are returned. From this resultant subset of
%   disparate data objects we create a single synthetic data product
%   that represents a 'materialization' of the observations. Technical
%   discussion will include high-level descriptions of the
%   infrastructure and tools employed with focus on our development in
%   the rapidly evolving semantic frontier where scalability and best
%   practices are particularly germane and equally elusive.
\end{abstract}

### 2011-06-05 docs/pubs/EIM-2011/application.tex
Trivial spelling and grammar fixes.

By leveraging the relationships defined and/or inferred from the
ontology we are able to increase recall beyond what is possible for
simple keyword-based searches
\cite{berkley09:_improv_data_discov_for_metad}. Broad queries return
direct matches and also n-depth subclass matches. The queries can be
quickly honed when using this browsing paradigm which allows rapid
exploration of the data repository without the oneous of defining
complete observational queries \emph{de novo}. Measurement ``templates''
as defined in OBOE compatible ontologies enable a single concept to
proxy its constituent parts, namely the characteristics of
particular entities that can be measured with a set of protocols and
standards. This short-hand query generation highlights a
compelling reason for using OBOE extension ontologies and can
also be leveraged when creating or editing semantic annotations in the Morpho interface.

@@

\mypara{Data integration.} The materialization routine used for semantic data queries laid the groundwork for our ongoing data integration pursuits. In addition to inspecting the data for values within a range and returning the datasets that contain a match, the ``smart'' data integration feature goes further by constructing a synthetic data product that represents the complete results of the query in terms of both the attributes and the values that are returned. Each original data object may have wildly different syntactic structures (e.g. column number, naming, order) but could share a subset of attributes that are semantically compatible as defined in accompanying annotations. These compatible, in-common attributes become the facets and metrics in a synthetic dataset; their values filtered accordingly (\figref{fig:integration}). While preliminary results have been quite promising - albeit academic - continued development of algorithms for determining compatibility of annotated measurements (e.g., gram and ounce are both mass units) and for mutating measurement values using ontologically-defined conversions (e.g., 1000 milligrams in a gram) will enable increasingly thorough, automated data integration practices.

>>>>

By leveraging the relationships defined and/or inferred from the
ontology we are able to increase recall beyond what is possible for
simple keyword-based searches
\cite{berkley09:_improv_data_discov_for_metad}. Broad queries return
direct matches and also n-depth subclass matches. The queries can be
quickly honed when using this browsing paradigm which allows rapid
exploration of the data repository without the onerous task of defining
complete observational queries \emph{de novo}. Measurement ``templates''
as defined in OBOE compatible ontologies enable a single concept to
proxy its constituent parts, namely the characteristics of
particular entities that can be measured with a set of protocols and
standards. This short-hand query generation highlights a
compelling reason for using OBOE extension ontologies and can
also be leveraged when creating or editing semantic annotations in the Morpho interface.

@@

\mypara{Data integration.} The materialization routine used for semantic data queries laid the groundwork for our ongoing data integration approach. In addition to inspecting the data for values within a range and returning the datasets that contain a match, the ``smart'' data integration feature goes further by constructing a synthetic data product that represents the complete results of the query in terms of both the attributes and the values that are returned. Each original data object may have wildly different syntactic structures (e.g. column number, naming, order) but could share a subset of attributes that are semantically compatible as defined in accompanying annotations. These compatible, in-common attributes become the facets and metrics in a synthetic dataset; their values filtered accordingly (\figref{fig:integration}). While preliminary results have been quite promising - albeit academic - continued development of algorithms for determining compatibility of annotated measurements (e.g., gram and ounce are both mass units) and for mutating measurement values using ontologically-defined conversions (e.g., 1000 milligrams in a gram) will enable increasingly thorough, automated data integration practices.

### 2011-06-05 docs/pubs/EIM-2011/intro.tex
Trivial spelling and grammar fixes.

A number of recent efforts within the earth and environmental
informatics communities are adopting the notion of an \emph{observation} as a
key modeling concept for enabling improved discovery and integration
of scientific data
\cite{om,fox09:_ontol,tarboton07:_cuahs_commun_obser_data_model,cushing07:_compon_based_end_user_datab,balhoff10:_phenex,bowers08}. These
approaches provide higher-level observational data models for
describing and representing observations and measurements found in
underlying datasets by defining common ``core'' concepts such as the
entities or features being observed, measurement units and protocols,
and context relationships between observations \cite{om,bowers08}.  A
major goal of these approaches is to enable interoperability and
uniform access to data by abstracting away the underlying
representation details that often exist across scientific data sets.

>>>>

A number of recent efforts within the earth and environmental
informatics communities are adopting the notion of an \emph{observation} as a
key modeling concept for enabling improved discovery and integration
of scientific data
\cite{om,fox09:_ontol,tarboton07:_cuahs_commun_obser_data_model,cushing07:_compon_based_end_user_datab,balhoff10:_phenex,bowers08}. These
approaches provide higher-level observational data models for
describing and representing observations and measurements found in
underlying datasets by defining common ``core'' concepts such as the
entities or features being observed, measurement units and protocols,
and context relationships between observations \cite{om,bowers08}.  A
major goal of these approaches is to enable interoperability and
uniform access to data by abstracting away the underlying
representation details that often impede integration across scientific data sets.
